---
title: "Exploratory Data Analysis of Offenses in Texas"
author: "Sayali Kambli"
date: "20/03/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(reshape2)
library(ggthemes)
library(hrbrthemes)
library(leaflet)
library(tidyverse) 
library(plotly) 
library(knitr)
library(kableExtra)
library(gridExtra)
#load the data
data <- read.csv("37-00049_UOF-P_2016_prepped.csv",header=TRUE)
data_df <- fortify(data[-1,])
```

## Index:  
  
1.0  Introduction

2.0  Analyses

     2.1 Analysis of Time of incidents          
     2.2 Types of Offenses    
     2.3 Distribution of Race of Officers and Offenders         
     2.4 Gender Distribution of Officers and Offenders     
     2.5 Relationship between Officer's experience and their violence  
     2.6 Types of Arrested offenders                             
     2.7 Variation of Race of Offenders in Texas
     2.8 Incident reasons in Texas 
     2.9 Variation of incidents in Year 2016 in Texas  
    
3.0  Conclusion

4.0  References

## 1.0. Introduction:

The data set that I am using is taken from  *https://www.kaggle.com/center-for-policing-equity/data-science-for-good*.
Name of the csv file is **37-00049_UOF-P_2016_prepped.csv**.
   

## 2.0. Analyses:   
 
 
### 2.1. Analysis of Time of incidents:
Dataset includes time of the incident.Below is the graph that shows top 30 incident times.

```{r a2, echo=FALSE}
attach(data_df)

#plot graph for time of incidents
time <- head(summary(as.factor(data_df$INCIDENT_TIME)),30)

Time_df <- data.frame(Time=as.vector(names(time)),Count=as.vector(as.integer(time)))
plot<- Time_df %>% ggplot(aes(x=reorder(Time,Count),y=Count))+geom_bar(stat="identity",fill="#a17c77")+
  theme_bw()+xlab("Time of day")+coord_flip()
 ggplotly(plot)
```
 As seen in the graph, most incidents were occurred at night time.
 
 
### 2.2. Types of Offenses
As I mentioned in my first Coursework, list of the 6 most common offenses occurred:

```{r Offense, echo=FALSE }

#Offense types

SUBJECTOFFENSE <- head(summary(as.factor(data_df$SUBJECT_OFFENSE)),6)

SUBJECTOFFENSE_df <- data.frame(SUBJECT_OFFENSE=as.vector(names(SUBJECTOFFENSE)),Count=as.vector(as.integer(SUBJECTOFFENSE)))

kable(SUBJECTOFFENSE_df) %>% kable_styling(position = "center")
 
```
### 2.3. Distribution of Race of Officers and Offenders
As I observed in my first coursework, officers and offenders had different race types.
```{r Race,echo=FALSE}
#Race Observation
SUBRACE <- ggplot(data_df, aes(SUBJECT_RACE))+ylab("Incidents")+
  theme_bw()+geom_bar(fill="#251e3b")+xlab("offenders Race")+ylab("Incident count")

OFFRACE <- ggplot(data_df, aes(OFFICER_RACE))+ylab("Incidents")+
  theme_bw()+geom_bar(fill="#251e3b")+xlab("Officer Race")+ylab("Incident count")

subplot(ggplotly(OFFRACE), ggplotly(SUBRACE), nrows=1,shareY =TRUE,titleX = TRUE)
```
In incidents White officers were more than other races where in case of Offenders Black was the most common race type.

### 2.4. Gender Distribution of Officers and Offenders 
 As I covered in my first coursework, following is the comparison of Genders in Officers and offenders:
```{r Gender ,echo=FALSE}
# Gender comparison
OFFGEN <- ggplot(data_df, aes(OFFICER_GENDER))+ylab("Incidents")+
  theme_bw()+ geom_bar(fill=c("#f75ee8", "#282ff7"))+xlab("Officer Gender")+ylab("Incident count")

SUBGEN <- ggplot(filter(data_df,SUBJECT_GENDER %in% c("Female","Male")), aes(SUBJECT_GENDER))+
  ylab("Incidents")+ 
  geom_bar(fill=c("#f75ee8", "#282ff7"))+xlab("offender Gender")+
  theme_bw()+ylab("Incident count")

subplot(ggplotly(OFFGEN), ggplotly(SUBGEN), nrows=1,shareY =TRUE,titleX = TRUE)
```
Male Officers/Offenders were way more than Female Officers/Offenders.It looks like Female Offenders were more than Female Officers.

### 2.5. Relationship between Officer's experience and their violence

Total applied force is the count of number of forces Officers used in incident( 0-10).
So,is there any relationship between experience of officers and their violence?
I have generated graph which will answer this question.
```{r violence,echo=FALSE}
 #Officer years on force vs applied force
data_df <- data_df %>%mutate(Total_force_Used=sapply(strsplit(data_df$FORCE_EFFECTIV, ","), length))
 graph2 <- ggplot(data_df,aes(y=as.numeric(OFFICER_YEARS_ON_FORCE),x=Total_force_Used))+geom_point()+theme_stata() + 
  scale_color_stata() +ylab("OFFICER YEARS ON FORCE")+xlab("Total Force Used")
 ggplotly(graph2)
```

As graph shows that Officers working for more years used less number of forces.
So it can be concluded that officers having less experience are usually more violent than the experienced officers in incidents.

### 2.6. Types of Arrested offenders
  Following are types of descriptions of offenders in arrests:
```{r description,echo=FALSE} 
#arrested subject's description
arrested_Data <- data_df %>% filter(SUBJECT_WAS_ARRESTED =="Yes")
graph3 <- ggplot(arrested_Data,aes(x=SUBJECT_DESCRIPTION))+coord_flip()+
  geom_bar(fill="#faf1aa")+theme_classic()+xlab("Offender DESCRIPTION")+ylab("Count of Arrest")
ggplotly(graph3)
```

As bar chart depicts, most times in arrests offender was either mentally unstable or on Alcohol or unknown drugs.

### 2.7. Variation of Race of Offenders in Texas
```{r map,echo=FALSE, warning=FALSE}
data1 <-data[-1,]
dataMAp<- data1 %>% mutate(lat=as.numeric(LOCATION_LATITUDE),long=as.numeric(LOCATION_LONGITUDE))  %>% filter(lat != 0 & long !=0)
colorselection <- colorFactor(palette =c("purple","maroon","blue","green","pink","Yellow","red"),dataMAp$SUBJECT_RACE)
dataMAp %>%  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(color = ~colorselection(SUBJECT_RACE),radius=2) %>%
  addLegend("bottomright", pal = colorselection, values = ~SUBJECT_RACE,
            title = "Subject Race")
```
As map shows,Black offenders were included in incidents more in South Texas where as White offenders were more in North part of Texas. 

### 2.8. Incident reasons in Texas

Following map shows incident reasons in Texas
```{r map2, echo=FALSE, warning=FALSE}
#map of incident reason

dataMAp1 <- dataMAp %>% filter(INCIDENT_REASON != "NULL")
colorselection <- colorFactor(palette =c("purple","black","blue","green","pink","Yellow","orange","maroon","white","red","grey","light blue","brown"), dataMAp1$INCIDENT_REASON)
dataMAp1 %>%  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(color = ~colorselection(INCIDENT_REASON),radius=2) %>%
  addLegend("bottomright", pal = colorselection, values = ~INCIDENT_REASON,
            title = "incident reasons")
```

As it is seen, most common reasons for incidents were Arrest and Service call and they are scattered all over Dallas in Texas

### 2.9. Variation of incidents in Year 2016 in Texas

Below is the incidents count occurred all over year 2016.

```{r year, echo=FALSE, warning=FALSE}
#Incident in Year 2016
data_year <- data_df%>%arrange(INCIDENT_DATE) %>% count(INCIDENT_DATE)  %>% mutate(INCIDENT_DATE=as.Date(INCIDENT_DATE,"%m/%d/%y"))
gbase <- data_year %>% ggplot(aes(x=as.Date(INCIDENT_DATE),y=n))+ geom_line(color="lightblue")+scale_x_date(limit=c(as.Date("2016-01-01"),as.Date("2016-12-31")),date_breaks = "2 month", date_labels = "%b")+
          xlab("Year 2016")+ylab("Incident count")+theme_stata()

ggplotly(gbase)
```
In 2016, incidents were less at end of the year compared with start or mid year.



## 3.0. Conclusion:

*  Most incidents were occurred at night time.
*  Male Officers/Offenders were way more than Female Officers/Offenders.
*  **White officers** and  **Black Offenders** were more than other race types.
*  officers having less experience were usually more violent than the experienced officers in incidents.
*  Offenders with  **mentally instability , Alcohol and unknown drugs** were more likely to get arrested.
*  Most common reasons for incidents were **Arrest and Service call** in Dallas.
*  Incidents in  **South Texas** had more **Black offenders** than other race types.
*  Incidents in  **North Texas** had more **White offenders** than other race types.
*  Incidents were less at end of the year 2016 compared with start or mid year.

## 4.0. References:

* Dataset: https://www.kaggle.com/center-for-policing-equity/data-science-for-good.
* CSV File : 37-00049_UOF-P_2016_prepped.csv

* 2.2. Types of Offenses : MA304 - Coursework 1 , Author- Sayali Kambli
* 2.3. Distribution of Race of Officers and Offenders : MA304 - Coursework 1 , Author- Sayali Kambli
* 2.4. Gender Distribution of Officers and Offenders  : MA304 - Coursework 1 , Author- Sayali Kambli