---
title: "Individual Project: Data analysis and statistics with R"
author: "Sayali Kambli"
date: "30/04/2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA)
library(ggplot2)
library(ggthemes)
library(plotly) 
library(knitr)
library(kableExtra)
library(gridExtra)
#load the data
data <- read.csv("mammals_sleep.csv",header = TRUE)
```

## Index:  
  
1.0  Introduction

2.0  Methods

     2.1 Cleaning Dataset          
     2.2 Analysing Sleep time density distribution     
     2.3 Analysing summary statistics of Remaining sleep and Awake time       
     2.4 Analysing Sleep time distribution over vores and conservation types 

3.0  Results

     3.1 Hypothesis Testing
     3.2 Linear Regression

4.0  Conclusion

5.0  References

## 1.0. Introduction:

The data set that I am using is taken from  *https://moodle.essex.ac.uk/course/view.php?id=15074&section=8*.
Name of the csv file is **mammals_sleep.csv**.
    
    This dataset has biological information  related to sleep time in a day on 83 mammals,
    which can be used to find  relation between the characteristics of mammals and their sleeping times.
    I am going to perform hypothesis testing and linear regression in this report.

## 2.0. Methods: 
### 2.1. Cleaning data
   
    Dataset has NA values which can mislead the result.
```{r clean, echo=FALSE }
#finding columns having NA values
coln <- colnames(data)[colSums(is.na(data)) > 0]

```
Following columns contain NA values:
 
*vore,conservation,sleep_rem,sleep_cycle and brainwt*
    
    I have removed this NA values as per the involved processes.
 
### 2.2. Analysing Sleep time density distribution 

Below graph shows Sleep time density distribution:

```{r density, echo=FALSE }
#Sleep time analyze
minSleep <-  filter(select(data,name,sleep_total),sleep_total==min(sleep_total))
maxSleep <-  filter(select(data,name,sleep_total),sleep_total==max(sleep_total))
meanSleep <-  summarise(data,meanSleep=mean(sleep_total))
ggplot(data,aes(x=sleep_total)) +geom_density() +
  labs(title="Total Sleep time density curve",x="Total Sleep(hours)", y = "Density") + 
  theme_bw()
 
```

   *  *``r minSleep[1]``* has minimum sleep hours of *``r round(minSleep[2],2)``*  whereas *``r maxSleep[1]``* sleeps for maximum hours of *``r round(maxSleep[2],2)``*.
   *  *``r round(meanSleep,2)``* hours is found to be average sleep time of mammals.
 
### 2.1. Analysing summary statistics of Remaining sleep and Awake time:

Following tables show summary statistics of remaining sleep time and awake time of mammals:

```{r stat, echo=FALSE}
#remaining sleep time has some NA so removing it
sleep_rem <-  na.omit(data$sleep_rem)
#summary of remaining sleep time
RemSleep <- data.frame(rbind("Remaining Sleep hours"=summary(sleep_rem)))
#summary of awake time
Awake <- data.frame(rbind("Awake hours"=summary(data$awake)))
#displaying data in table
kable(RemSleep)%>% kable_styling(position = "center",latex_options =c("hold_position"))
kable(Awake)%>% kable_styling(position = "center",latex_options =c("hold_position"))

```


 
### 2.2.Analysing Sleep time distribution over vores and conservation types

There are different types of vores and conservation types present in dataset.
Given graph shows sleep time distribution for these types

```{r voreconser, echo=FALSE }
#omitting NA from vore and conservation
datavoreconser <- filter(select(data,vore,sleep_total,conservation),vore != "NA" & conservation != "NA")

vore <- ggplot(datavoreconser, aes(y=sleep_total, x=vore,color=vore))+geom_point()+ labs(x = "vore", y = "Total Sleep",title = "Total sleep hours for vore")

con<-ggplot(datavoreconser, aes(y=sleep_total, x=conservation,color=conservation))+geom_point()+ labs(x = "Conservation", y = "Total Sleep",title = "Total sleep hours for conservation") 
#plot graphs 
grid.arrange(vore,con, ncol=1)
 
```

It can be seen from graphs that Sleeping time of Carnivores and LC(Least concern) conservation type mammals are well distributed over range of 0-20 hours. 

## 3.0. Results: 

### 3.1. Hypothesis Testing

I have performed hypothesis testing wrt Sleep time and vores of mammals.

  *H0: Mean sleep time is equal for two vores(omni and carni)*

```{r Hypothesis, echo=FALSE, warning=FALSE}
#omitting NA from vore for this process
datavore <- filter(select(data,vore,sleep_total),vore != "NA")

#hypothesis testing

m <- datavore %>% filter(vore %in% c("omni", "carni"))
omni <- (m %>% filter(vore =="omni"))$sleep_total
carni <- (m %>% filter(vore =="carni"))$sleep_total
#performing t test
testresult<- t.test(omni,carni,alternative = "two.sided")
#display testresult
testresult

```
As per the test result above, there is evidence against H0(p-value is relatively higher) so *Null hypothesis saying(Mean sleep time is equal for two vores(omni and carni)) can not be rejected.*

### 3.2. Linear Regression 

 Linear Regression between Total Sleep and (brain weight and body weight):
   I have taken Log of Brain weight and Body weight values.
```{r Covar, echo=FALSE, warning=FALSE, fig.height=6}
#omitting NA only from brainwt as only it has NA value from set below
data_df <- filter(select(data,sleep_total,brainwt,bodywt),brainwt != "NA")
#performing cor test to find covariance
corr_body_sleep <- cor.test(log(data_df$bodywt),data_df$sleep_total, method="pearson")

corr_brain_sleep <- cor.test(log(data_df$brainwt),data_df$sleep_total, method="pearson")  
```
  First of all Checking covariance between them:
  
     1. Covariance between Total Sleep and Body Weight is `r corr_body_sleep$estimate`
     2. Covariance between Total Sleep and Brain Weight is `r corr_brain_sleep$estimate`
    
    Both covariance values are relatively strong .So Linear regression can be
    performed here.(Negative values are anti-correlation, sleep time is more than body or brain weight)


Now performing Linear Regression,


```{r LR, echo=FALSE, warning=FALSE, fig.height=6}


#plotting graph with Linear Regression
plotbodywt<- ggplot(data_df, aes(x=sleep_total, y= log(bodywt))) +
  geom_point(col=1, size=1)  + labs(x="Total Sleep",y="Log(body weight)") +
  geom_smooth(method = "lm", se=FALSE, color="brown", formula = y ~ x)  +theme_stata()

plotbrainwt <- ggplot(data_df, aes(x=sleep_total, y= log(brainwt))) +
  geom_point(col=1, size=1)  + labs(x="Total Sleep",y="Log(brain weight)") +
  geom_smooth(method = "lm", se=FALSE, color="brown", formula = y ~ x) +theme_stata()
#plot graph
grid.arrange(plotbodywt,plotbrainwt, ncol=1)

#Calculating MAE 
linearModelbodysleep <- lm(sleep_total ~ log(bodywt), data=data_df)
linearModelbrainsleep <- lm(sleep_total ~ log(brainwt), data=data_df)
MAEbodysleep <- mean(abs(linearModelbodysleep$residuals))
MAEbrainsleep <- mean(abs(linearModelbrainsleep$residuals))
```



#### Calculating MAE for models     

* MAE for Linear regression model between Total Sleep and Body Weight is *``r round(MAEbodysleep,2) ``*
* MAE for Linear regression model between Total Sleep and Brain Weight is *``r round(MAEbrainsleep,2)``*

    Both MAE values are relatively smallest which indicates that Linear regression model 
    is good fit for this variables.

## 4.0. Conclusion:

*  *``r minSleep[1]``* has minimum sleep hours of *``r round(minSleep[2],2)``*.
*  *``r maxSleep[1]``* sleeps for maximum hours of *``r round(maxSleep[2],2)``*.
*  *``r round(meanSleep,2)``* is found to be average sleep time of mammals.
*  *``r RemSleep$Min``* hours is minimum remaining sleep  whereas *``r RemSleep$Max``* hours is maximum remaining sleep.
*  *``r round(RemSleep$Mean,2)``* hours is the average remaining sleep time of mammals.
*  Awake time for mammals ranges from *``r Awake$Min``* hours to  *``r Awake$Max``* hours ,having mean time of *``r round(Awake$Mean,2)``* hours.
*  Sleeping time of *Carnivores and LC(Least concern)* mammals are well distributed over range of 0-20 hours
*  Mean sleep time is relatively similar for two vores(omni and carni) of mammals.
*  *Linear regression* model is good fit for Total Sleep and (brain weight and body weight).


## 5.0. References:

* Dataset: https://moodle.essex.ac.uk/mod/resource/view.php?id=805364.
* CSV File : mammals_sleep.csv
